---
title: "Chapter 2 - Small Worlds and Large Worlds"
subtitle: "Chapter code and examples"
author: "Tiago"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
library(tidyverse)
```

## Section 2.4.3 - Grid Approximation

### Building a grid approximation in R for the globe tossing problem 

The goal is to compute the posterior probability of probability of our right 
index finger water lands in water when we throw the globe to air. 

The first nine samples are 

$$
W \ L \ W \ W \ W \ L \ W \ L \ W
$$

So in our data we landed our right index on water $6$ times in $9$ throws. 

### Model specification 

$$
W \sim Binomial(N, p)
$$
$$
p \sim Uniform(0,1)
$$
Since, $p$ has uniform prior between $0$ and $1$, then 

$$
Pr(p) = \frac{1}{1-0} = 1
$$

### Grid approximation of the posterior distribution

The recipe is the following

1. Define a grid with the number of points I want to estimate the posterior;

2. Compute the prior at each parameter value on the grid;

3. Compute the likelihood at each parameter value

4. Compute the unstandardized posterior values by multiplying 
`likelihood * prior`.

5. Standardize the posterior distribution by multiplying by the sum of all values.

Here it is what it looks like in R

```{r}
# number of grid points
n_points <- 20

# Define a grid
p_grid <- seq( from = 0, to  = 1, length.out = n_points)

# define prior
prior <- rep(1, n_points)

# compute likelihood at each grid point
likelihood <- dbinom(6, size = 9, prob = p_grid)

# compute joing probability of data and parameters
# that is, the product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardize the posterior, so it sums to one
posterior <- unstd.posterior / sum(unstd.posterior)

# save grid and posterior into a data frame
dgrid <- 
  tibble(
    p_grid, 
    posterior
  )
```

#### Ploting posterior grid approximation of $p$

```{r}
dgrid %>% 
  ggplot(aes(x = p_grid, y = posterior)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Posterior approximation of p using 20 grid points"
  )
```

### We can try with different numbers of grid points

```{r}
n_points <- c(3,10, 20, 100, 400, 1000) 

dgrid <- map_dfr(n_points, function(points){
  
  # Define a grid
  p_grid <- seq( from = 0, to  = 1, length.out = points)
  
  # define prior
  prior <- rep(1, points)
  
  # compute likelihood at each grid point
  likelihood <- dbinom(6, size = 9, prob = p_grid)
  
  # compute joing probability of data and parameters
  # that is, the product of likelihood and prior
  unstd.posterior <- likelihood * prior
  
  # standardize the posterior, so it sums to one
  posterior <- unstd.posterior / sum(unstd.posterior)
  
  # save grid and posterior into a data frame
  dgrid <- 
    tibble(
      grid_points = as.character(points),
      p_grid, 
      posterior
    ) 
  return(dgrid)
}) %>% 
  mutate(grid_points = factor(grid_points, levels = n_points))
```

#### Ploting the posteriro approximation of $p$

```{r}
dgrid %>% 
  ggplot(aes(x = p_grid, y = posterior)) +
  geom_line() +
  geom_point(aes(color = grid_points)) +
  labs(
    title = "Posterior grid approximations with varying numbers of grid points"
  ) +
  facet_wrap(~ grid_points, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none")
  
```

### Now with different priors

#### Step function prior

```{r}
step_prior <- function(p_grid) { 
  prior <- ifelse(p_grid < 0.5, 0, 1)
  return(prior)
  }
```

```{r echo=FALSE}
n_points <- 50

# Define a grid
p_grid <- seq( from = 0, to  = 1, length.out = n_points)

tibble(
  p_grid
) %>% 
  mutate(prior = step_prior(p_grid)) %>% 
  ggplot(aes(x = p_grid, y = prior)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Prior probability of parameter p"
  )

```

#### Approximating the posterior

```{r}
# number of grid points
n_points <- 20

# Define a grid
p_grid <- seq( from = 0, to  = 1, length.out = n_points)

# define prior
prior <- step_prior(p_grid)

# compute likelihood at each grid point
likelihood <- dbinom(6, size = 9, prob = p_grid)

# compute joing probability of data and parameters
# that is, the product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardize the posterior, so it sums to one
posterior <- unstd.posterior / sum(unstd.posterior)

# save grid and posterior into a data frame
dgrid <- 
  tibble(
    p_grid, 
    posterior
  )

dgrid %>% 
  ggplot(aes(x = p_grid, y = posterior)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Posterior approximation of p using 20 grid points"
  )
```

#### Double exponential prior

```{r}
dbl_exp <- function(p_grid) { 
  prior <- exp(-5*abs(p_grid - .5))
  return(prior)
  }
```

```{r echo=FALSE}
n_points <- 50

# Define a grid
p_grid <- seq( from = 0, to  = 1, length.out = n_points)

tibble(
  p_grid
) %>% 
  mutate(prior = dbl_exp(p_grid)) %>% 
  ggplot(aes(x = p_grid, y = prior)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Prior probability of parameter p"
  )

```

#### Approximating the posterior

```{r}
# number of grid points
n_points <- 20

# Define a grid
p_grid <- seq( from = 0, to  = 1, length.out = n_points)

# define prior
prior <- dbl_exp(p_grid)

# compute likelihood at each grid point
likelihood <- dbinom(6, size = 9, prob = p_grid)

# compute joing probability of data and parameters
# that is, the product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardize the posterior, so it sums to one
posterior <- unstd.posterior / sum(unstd.posterior)

# save grid and posterior into a data frame
dgrid <- 
  tibble(
    p_grid, 
    posterior
  )

dgrid %>% 
  ggplot(aes(x = p_grid, y = posterior)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Posterior approximation of p using 20 grid points"
  )


```



